# Graphing

```{r, echo=FALSE}
# Unattach any packages that happen to already be loaded. In general this is unnecessary
# but is important for the creation of the book to not have package namespaces
# fighting unexpectedly.
pkgs = names(sessionInfo()$otherPkgs)
if( length(pkgs > 0)){
  pkgs = paste('package:', pkgs, sep = "")
  for( i in 1:length(pkgs)){
    detach(pkgs[i], character.only = TRUE, force=TRUE)
  }
}
knitr::opts_chunk$set(cache=FALSE)
```

In this chapter you will learn more about the `ggplot2` package, part of the `tidyverse` suite of packages.

```{r, eval=FALSE}
library(ggplot2)
```

```{r, message=FALSE, warning=FALSE, results='hide', include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)   # loading ggplot2 and dplyr
})
```

Dr. Sonderegger's Video Companion: [Video Lecture](https://youtu.be/eJT9EdiWlEg).

## Introduction to ggplot2 graphics

There are three major “systems” of making graphs in R. The basic plotting commands in R are quite effective but the commands do not have a way of being combined in easy ways, often requiring loops or functions to produce complex multi-layered graphs. Lattice graphics (which the `mosaic` package uses) makes it possible to create some quite complicated graphs but it is very difficult to make non-standard graphs. The package we will focus on here, `ggplot2`, tries to not anticipate what the user wants to do, but rather provide the mechanisms for pulling together different graphical concepts and the user gets to decide which elements to combine. This will allow a user to create almost any graphic of interest. 

To make the most of `ggplot2` it is important to wrap your mind around “The Grammar of Graphics”. Briefly, the act of building a graph can be broken down into three steps. 

1. Define what data set we are using. 

2. What is the major relationship we wish to examine? 

3. In what way should we present that relationship? These relationships can be presented in multiple ways, and the process of creating a good graph relies on building layers upon layers of information. 

For example, after preparing a proper data frame for the analysis, we may want to present a linear model result including both the data, best-fit line, and other statistical intervals (confidence, prediction). We might start with printing the raw data as a scatter graphic. We can then add additional layers such as an overlay of the regression line over the top of the scatter. We can then continue this thinking for as many layers as we want, producing a final graphic that conveys important information clearly. 

By producing code that outputs these graphics, we can then introduce small changes as needed to improve them as our needs grow. Although the first version of the graph may take some time to prepare, once code is stored in a script, it becomes easy to update and improve the graphic over time. This is an essential part of work that is done in research laboratories or for publications, and being able to easily handle the updates needed to a graphic can not only save time, but help you prepare the exact graphic needed to tell your story! 

Next, it should be noted that `ggplot2` is designed to act on data frames. It is actually hard to just draw three data points and for simple graphs it might be easier to use the base graphing system in R. However for any real data analysis project, the data will already be in a data frame and this is not an annoyance. Throughout this textbook, you will become very comfortable with formatting data into data frames.

These notes are sufficient for creating simple graphs using `ggplot2`, but are not intended to be exhaustive. There are many places online to get help with `ggplot2`. One very nice resource is the website, [http://www.cookbook-r.com/Graphs/](http://www.cookbook-r.com/Graphs/), which gives much of the information available in the textbook `R Graphics Cookbook` which I highly recommend. Second is just using Google to solve your problems and see what you can find on websites such as StackExchange. As you become more advanced in your knowledge of preparing figures through `ggplot2`, generative AI sources such as chatGPT can help produce the syntax outline that you can then manipulate to create the exact graphic of interest.

One way that `ggplot2` makes it easy to form very complicated graphs is that it provides a large number of basic building blocks that, when stacked upon each other, can produce extremely complicated graphs. A full list is available at [https://ggplot2.tidyverse.org/reference/](https://ggplot2.tidyverse.org/reference/). The following list gives some idea of different building blocks that are most commonly used. These different geometries offer different ways to display the relationship between variables and can be combined in many interesting ways. We will explore several geometries in this chapter and future graphing chapters.

|  Geom	           |    Description                                     |   Required Aesthetics  |
|------------------|----------------------------------------------------|------------------------|
| `geom_histogram` | A histogram                                        | `x`    |
| `geom_bar`       | A barplot (y is number of rows)                    | `x`    |
| `geom_col`       | A barplot (y is given by a column)                 | `x, y` |
| `geom_density`   | A density plot of data. (smoothed histogram)       | `x`    |
| `geom_boxplot`   | Boxplots                                           | `x, y` |
| `geom_line`      | Draw a line (after sorting x-values)               | `x, y` |
| `geom_path`      | Draw a line (without sorting x-values)             | `x, y` |
| `geom_point`     | Draw points (for a scatterplot)                    | `x, y`                 |
| `geom_smooth`	   |  Add a ribbon that summarizes a scatterplot        | `x, y`                 |
| `geom_ribbon`	   | Enclose a region, and color the interior           | `ymin, ymax` |
| `geom_errorbar`  | Error bars                                         | `ymin, ymax` |
| `geom_text`      |  Add text to a graph (with box)                    | `x, y, label` |
| `geom_label`     | Add text to a graph  (without box)                 | `x, y, label`          |
| `geom_tile`      |  Create Heat map                                   | `x, y, fill`           |


A graph can be built up layer by layer, where:

* Each layer corresponds to a geometry (`geom`), each of which requires a data set and a mapping between an aesthetic (`aes`) and a column of the data set.
    * If you don't specify the data or aesthetic within a geometry, then the layer inherits everything defined in the primary `ggplot()` command.
    * You can have different data sets for each layer!
* Layers can be added with a `+`, or you can define two plots and add them together (second one over-writes anything that conflicts).

## Basic Graphs

### Scatterplots

To start with, we'll make a very simple scatterplot using the `iris` data set. The `iris` data set contains observations on three species of iris plants where we've measured the length and width of both the petals and sepals.  We will make a scatterplot of `Sepal.Length` versus `Petal.Length`, which are two columns in the data set. First, lets recall how to load a data set into our local environment. We then show the structure (`str`) of the data set to understand its data types and get a clear listing of the variables present in the data frame

```{r}
data(iris)  # load the iris data set that comes with R
str(iris)   # what columns do we have to play with...
```

We then make our first graphic using a single line of ggplot2 syntax.

```{r, fig.height=3, warning=FALSE, message=FALSE}
ggplot( data=iris, aes(x=Sepal.Length, y=Petal.Length) ) +  
	geom_point(  ) 
```

1. All ggplot graphics start with the `ggplot()` command.
2. The data set we wish to use is specified using `data=iris`.
3. The relationship we want to explore is `x=Sepal.Length` and `y=Petal.Length`. This means the x-axis will be the Sepal Length and the y-axis will be the Petal Length. To use data directly from the initial data frame, we must always include this mapping within the `aes()` command.
4. The way we want to display this relationship is through graphing 1 point for every observation (scatter). Thus  we map this information into the scatter graph geometry known as `geom_point()`.

We can define other attributes that might reflect other aspects of the data. For example, we might want for the color of the data point to change dynamically based on the species of iris. We can update our graph by including a second aesthetic (`aes()`) within the `geom_point()` specifically. Below we add a color option based on the species of iris. Notice below that by coloring the graph by species, we gain significant new information regarding how differences between the iris species.

```{r, fig.height=3}
ggplot( data=iris, aes(x=Sepal.Length, y=Petal.Length) ) +  
	geom_point( aes(color=Species) )
```

Let's take a moment to think about the `aes()` command inside the previous section of code, which can be quite mysterious to new users. The way to think about the `aes()` is that it gives you a way to define relationships that are dependent on the data frame (here `data=iris`). In the previous graph, the x-value and y-value for each point was defined dynamically by the data, as was the color. If we just wanted all the data points to be colored blue and larger, then the following code would do that

```{r, fig.height=3}
ggplot( data=iris, aes(x=Sepal.Length, y=Petal.Length) ) +  
	geom_point( color='blue', size=4 )
```

The important part isn't that color and size were defined in the `geom_point()` but that they were defined outside of an `aes()` function! Notice though by removing the dependence to the species, we have lost the ability to see how different each species behaves!

A few quick notes on the use of aesthetics:

1. Anything set inside an `aes()` command will be of the form `attribute=Column_Name` and will change based on the data. We must always use the `aes()` command to extract information from the data frame.
2. Anything set outside an `aes()` command will be in the form `attribute=value` and will be fixed. This is used for static information that doesn't depend on the data frame. 

### Box Plots 

Boxplots are a common way to show a categorical variable on the x-axis and continuous on the y-axis. Lets evaluate the distribution of `Petal.Length` versus the `Species` of iris. The syntax works very similar to above, but now we will map the information into a boxplot geometry.

```{r}
ggplot(iris, aes(x=Species, y=Petal.Length)) +
  geom_boxplot()  
```

The boxes show the $25^{th}$, $50^{th}$, and $75^{th}$ percentile and the lines coming off the box extend to the smallest and largest non-outlier observation. Boxplots are important for viewing information across different categorical groupings, and are often related to ANOVA-type statistical analyses.  We will explore boxplots further in later chapters, as they are very common in all fields!

## Faceting

The goal with faceting is to make many panels of graphics where each panel represents the same relationship between variables, but something changes between each panel. For example using the `iris` data set we could look at the relationship between `Sepal.Length` and `Petal.Length` again, but now instead of placing all information with color on a single graph, we can make one panel per species. Faceting allows for the introduction of even more variables within a single graph!

```{r, fig.height=3}
library(ggplot2)
ggplot(iris, aes(x=Sepal.Length, y=Petal.Length)) +
  geom_point() +
  # facet_grid( cols = vars(Species) ) # using vars() to dictate columns
  # facet_grid( rows = vars(Species) ) #using vars() to dictate rows
  facet_grid( . ~ Species ) # Formula notation y ~ x
```

The line `facet_grid( formula )` tells `ggplot2` to make panels, and the formula tells how to orient the panels. In R, formulas are always interpreted in the order `y ~ x`. Because I want the species to change as we go across a page (horizontally, commonly called the x-direction on a 2D-graph), but don't have anything I want to change vertically (y-direction) we use `. ~ Species` to represent that. If we had wanted three graphs stacked vertically instead, we could use `Species ~ .`. Additional code is provided that allows the user to more specifically define columns (`cols = vars(Species)`). This syntax provides more control over the x- and y-direction, but the formula notation should become more familiar with experience in R.

For a second example, we look at a data set that examines the amount a waiter was tipped by 244 different parties. Covariates that were measured include the day of the week, size of the party, total amount of the bill, amount tipped, whether there were smokers in the group and the gender of the person paying the bill. Like always, lets load the data and explore the variables a bit, here we output the first 6 rows using the `head()` function.

```{r}
data(tips, package='reshape')
head(tips)
```

It is easy to look at the relationship between the size of the bill and the percent tipped. We use the flexible syntax of R to calculate the percent tip directly within the `aes()` command. We then plot the total bill against the tip/total_bill ratio (or percent tipped).

```{r, fig.height=3}
ggplot(tips, aes(x = total_bill, y = tip / total_bill )) +
  geom_point()
```

Next we investigate if there is a difference in tipping percent based on gender or day of the week by plotting this relationship for each combination of gender and day.

```{r, fig.height=3}
ggplot(tips, aes(x = total_bill, y = tip / total_bill, color=time )) +
  geom_point( ) +
  # facet_grid( day ~ sex )  # changing orientation emphasizes certain comparisons!
  facet_grid( sex ~ day )
```

It doesn't seem there is much difference between the genders, but we would like to investigate the relation to day of week further. Sometimes we want multiple rows and columns of the facets, but there is only one categorical variable with many levels. Rather than making the graph grow very long either horizontally or vertically, we use `facet_wrap()` which takes a one-sided formula and will wrap in both the x- and y-directions automatically.

```{r, fig.height=3}
ggplot(tips, aes(x = total_bill, y = tip / total_bill )) +
  geom_point() +
  # facet_grid( . ~ day)  # Four graphs in a row, Too Squished left/right!
  facet_wrap( ~ day )   # spread graphs out both left/right and up/down.
```

Finally we can allow the x and y scales to vary between the panels by setting “free”, “free_x”, or “free_y”. In the following code, the y-axis scale changes for each day in the data set.

```{r, fig.height=3}
ggplot(tips, aes(x = total_bill, y = tip / total_bill )) +
  geom_point() +
  facet_wrap( . ~ day, scales="free_y" )
```

## Annotation 

### Axis Labels and Titles

To make a graph more understandable, it is necessary to tweak the axis labels and add a main title and such. Lets return to the iris data set example, but improve the quality of our scatter graph. The code below introduces lots of new syntax that changes the written text in many areas of the graph. You could either call the `labs()` command repeatedly with each label, or you could provide multiple arguments to just one `labs()` call.  For readability, the lines have been separated in the code below.

```{r, fig.height=3}
# Assign the graph to the object 'p' allowing us to add more later.
P <-
  ggplot( data=iris, aes(x=Sepal.Length, y=Petal.Length, color=Species) ) +  
  geom_point( ) +
  labs( title='Sepal Length vs Petal Length' ) +
  labs( x="Sepal Length (cm)", y="Petal Length (cm)" ) +
  labs( color="Species Name")  + 
  labs( caption = "Iris data from Edgar Anderson (1935)" )

# Print out the plot
P
```

### Text Labels

One way to improve the clarity of a graph is to remove the legend and, instead, label the points directly on the graph. For example, we could instead have the species names near the cloud of data points for the species. 

Usually our annotations aren't stored in the data frame that contains our data of interest. So we need to either create a new (usually small) data frame that contains all the information needed to create the annotation or we need to provide all necessary information within the syntax. It is encouraged to make additional, smaller data frames, that store auxiliary information such as annotations as these can be easily changed without interfering with the syntax making the graphs. Either way, we need to specify the `x` and `y` coordinates, and the `label` to be printed, as well as any other attribute that is set in the global `aes()` command. That means if `color` has been set globally, the annotation layer also needs to address the `color` attribute.

#### Using a data frame

To do this in ggplot, we need to make a data frame that has the columns `Sepal.Length` and `Petal.Length` so that we can specify where each label should go, as well as the label that we want to print. This is essentially indicating where on the x-y graph the text should be located. Also, because color is matched to the `Species` column, this small data set should also have a the `Species` column.

This step always requires a bit of fussing with the graph because the text size and location should be chosen based on the size of the output graphic and if I rescale the image it often looks awkward. Typically I leave this step until the figure is being prepared for final publication.

```{r}
# create another data frame that has the text labels I want to add to the graph.
annotation.data <- data.frame(
  Sepal.Length = c(4.5,  6.5,  7.0),  # Figured out the label location by eye.
  Petal.Length = c(2.25, 3.75, 6.5),  # If I rescale the graph, I would redo this step.
  Species = c('setosa', 'versicolor', 'virginica'),
  Text = c('SETOSA', 'VERSICOLOR', 'VIRGINICA')
)

# Use the previous plot I created, along with the 
# aes() options already defined.
P +
  geom_text( data=annotation.data, aes(label=Text), size=2.5) +   # write the labels
  theme( legend.position = 'none' )                               # remove the legend

```


#### Setting attributes in-line

Instead of creating a new data frame, we could just add a new layer and  set all of the graph attributes manually. To do this, we have to have one layer for each text we want to add to the graph. The `annotate()` function takes a geometry layer type, here we use the `geom_text` layer to place the information on the graph. We must provide all necessary inputs for each line, but by doing this work manual we can avoid the annoyance of building a data frame for the label information.

```{r}
P +
  annotate('text', x=4.5, y=2.25, size=6, color='#F8766D', label='SETOSA'     ) +
  annotate('text', x=6.5, y=3.75, size=6, color='#00BA38', label='VERSICOLOR' ) +
  annotate('text', x=7.0, y=6.50, size=6, color='#619CFF', label='VIRGINICA'  ) +
  theme(legend.position = 'none')
```

Finally there is a `geom_label` layer that draws a nice box around what you want to print.

```{r}
P +
  annotate('label', x=4.5, y=2.25, size=6, color='#F8766D', label='SETOSA'     ) +
  annotate('label', x=6.5, y=3.50, size=6, color='#00BA38', label='VERSICOLOR' ) +
  annotate('label', x=7.0, y=6.75, size=6, color='#619CFF', label='VIRGINICA'  ) +
  theme(legend.position = 'none')
```

My recommendation is to just set the `x`, `y`, and `label` attributes manually inside an `annotate()` call if you have one or two annotations to print on the graph. If you have many annotations to print, then create a data frame that contains all of the required information and use `data=` argument.

## Exercises  {#Exercises_IntroGraphing}

### Exercise 1 {-}

Examine the data set `trees`, which should already be pre-loaded. Look at the help file using `?trees` for more information about this data set. We wish to build a scatter plot that compares the height and girth of these cherry trees to the volume of lumber that was produced.

**a)** Create a graph using `ggplot2` with Height on the x-axis, Volume on the y-axis, and Girth as the either the size of the data point or the color of the data point. Which do you think is a more intuitive representation?

**b)** Add appropriate labels for the main title and the x and y axes.

**c)** The R-squared value for a regression through these points is 0.36 and the p-value for the statistical significance of height is 0.00038. Add text labels "R-squared = 0.36" and "p-value = 0.0004" somewhere *on* the graph. 
    
### Exercise 2 {-}

Consider the following small data set that represents the number of times per day my wife played "Ring around the Rosy” with my daughter relative to the number of days since she has learned this game. The column `yhat` represents the best fitting line through the data, and `lwr` and `upr` represent a 95% confidence interval for the predicted value on that day. *Because these questions ask you to produce several graphs and evaluate which is better and why, please include each graph and response with each sub-question.*

```{r}
Rosy <- data.frame(
  times = c(15, 11, 9, 12, 5, 2, 3),
  day   = 1:7,
  yhat  = c(14.36, 12.29, 10.21, 8.14, 6.07, 4.00,  1.93),
  lwr   = c( 9.54,  8.5,   7.22, 5.47, 3.08, 0.22, -2.89),
  upr   = c(19.18, 16.07, 13.2, 10.82, 9.06, 7.78,  6.75))
```

**a)** Using `ggplot()` and `geom_point()`, create a scatterplot with `day` along the x-axis and `times` along the y-axis.

**b)** Add a line to the graph where the x-values are the `day` values but now the y-values are the predicted values which we've called `yhat`. Notice that you have to set the aesthetic `y=times` for the points and `y=yhat` for the line. Because each `geom_` will accept an `aes()` command, you can specify the `y` attribute to be different for different layers of the graph.

**c)** Add a ribbon that represents the confidence region of the regression line. The `geom_ribbon()` function requires an `x`, `ymin`, and `ymax` columns to be defined. For examples of using `geom_ribbon()` see the online documentation: [https://ggplot2.tidyverse.org/reference/geom_ribbon.html](https://ggplot2.tidyverse.org/reference/geom_ribbon.html).

```{r, eval=FALSE}
ggplot(Rosy, aes(x=day)) +
  geom_point(aes(y=times)) +
  geom_line( aes(y=yhat)) +
  geom_ribbon( aes(ymin=lwr, ymax=upr), fill='salmon')
```

**d)** What happened when you added the ribbon? Did some points get hidden? If so, why?

**e)** Reorder the statements that created the graph so that the ribbon is on the bottom and the data points are on top and the regression line is visible.

**f)** The color of the ribbon fill is ugly. Use Google to find a list of named colors available to `ggplot2`. For example, I googled “ggplot2 named colors” and found the following link: [http://sape.inf.usi.ch/quick-reference/ggplot2/colour](http://sape.inf.usi.ch/quick-reference/ggplot2/colour). Choose a color for the fill that is pleasing to you.

**g)** Add labels for the x-axis and y-axis that are appropriate along with a main title.

### Exercise 3 {-}

We'll next make some density plots that relate several factors towards the birth weight of a child. *Because these questions ask you to produce several graphs and evaluate which is better and why, please include each graph and response with each sub-question.*

**a)** The `MASS` package contains a data set called `birthwt` which contains information about 189 babies and their mothers. In particular there are columns for the mother's race and smoking status during the pregnancy. Load the `birthwt` by either using the `data()` command or loading the `MASS` library.

**b)** Read the help file for the data set using `MASS::birthwt`. The covariates `race` and `smoke` are not stored in a user friendly manner. For example, smoking status is labeled using a 0 or a 1. Because it is not obvious which should represent that the mother smoked, we'll add better labels to the `race` and `smoke` variables. For more information about dealing with factors and their levels, see the `Factors` chapter in these notes.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
data('birthwt', package='MASS')
birthwt <- birthwt %>% mutate(
  race  = factor(race,  labels=c('White','Black','Other')),
  smoke = factor(smoke, labels=c('No Smoke', 'Smoke')))
```
        
**c)** Graph a histogram of the birth weights `bwt` using `ggplot(birthwt, aes(x=bwt)) + geom_histogram()`.

**d)** Make separate graphs that denote whether a mother smoked during pregnancy by appending `+ facet_grid()` command to your original graphing command.

**e)** Perhaps race matters in relation to smoking. Make our grid of graphs vary with smoking status changing vertically, and race changing horizontally (that is the formula in `facet_grid()` should have smoking be the y variable and race as the x).

**f)** Remove `race` from the facet grid, (so go back to the graph you had in part d). I'd like to next add an estimated density line to the graphs, but to do that, I need to first change the y-axis to be density (instead of counts), which we do by using `aes(y=..density..)` in the `ggplot()` aesthetics command.

**g)** Next we can add the estimated smooth density using the `geom_density()` command.

**h)** To really make this look nice, lets change the fill color of the histograms to be something less dark, lets use `fill='cornsilk'` and `color='grey60'`. To play with different colors that have names, check out the following: [https://www.datanovia.com/en/blog/awesome-list-of-657-r-color-names/](https://www.datanovia.com/en/blog/awesome-list-of-657-r-color-names/).

**i)** Change the order in which the histogram and the density line are added to the plot. Does it matter and which do you prefer?

**j)** Finally consider if you should have the histograms side-by-side or one on top of the other (i.e. `. ~ smoke` or `smoke ~ .`). Which do you think better displays the decrease in mean birth weight and why?

### Exercsie 4 {-}

Load the data set `ChickWeight`, which comes pre-loaded in R, and get the background on the data set by reading the manual page `?ChickWeight`. *Because these questions ask you to produce several graphs and evaluate which is better and why, please include each graph and response with each sub-question.*

**a)** Produce a separate scatter plot of weight vs age for each chick. Use color to distinguish the four different `Diet` treatments. *Note, this question should produce 50 separate graphs! If the graphs are too squished you should consider how to arrange them so that the graphs wrap to a new row of graphs in the resulting output figure. The results are messy!*

**b)** We could examine these data by producing a scatter plot for each diet. Most of the code below is readable, but if we don't add the `group` aesthetic the lines would not connect the dots for each Chick but would instead connect the dots across different chicks.

```{r, eval=FALSE}
data(ChickWeight)
ggplot(ChickWeight, aes(x=Time, y=weight, group=Chick )) +
  geom_point() + geom_line() +
  facet_grid( ~ Diet) 
```
        
*Notice in the code chunk above, if you copied from the online source code you must remove the `eval=FALSE` in the chunk header. This option allows the code to be displayed, but it won't be run and no plot will be produced in your final output document. So when you ask, why don't I see a plot?, I'll reminder you of this statement!*
