---
title: "Assignment 6"
author: "Your Name Here"
date: "`r Sys.Date()`"
output: pdf_document
---



## Directions:
This assignment covers chapter 6 and  chapter 7.  Please show all work in this document and knit your final draft into a pdf.  Factors are what R calls categorical variables.  Factors can be qualitative or a quantitative variable that we have chopped into a small number of non-overlapping bins or a few discrete values.

## Problem 1: For Loops

I often want to repeat some section of code some number of times. For example, I might want to create a bunch plots that compare the density of a t-distribution with specified degrees of freedom to a standard normal distribution. 
    
```{r, fig.height=3, message=FALSE, warning=FALSE}
    library(ggplot2)
    df <- 5
    N <- 1000
    x.grid <- seq(-3, 3, length=N)
    data <- data.frame( 
      x = c(x.grid, x.grid),
      y = c(dnorm(x.grid), dt(x.grid, df)),
      type = c( rep('Normal',N), rep('T',N) ) )
    
    # make a nice graph
    myplot <- ggplot(data, aes(x=x, y=y, color=type, linetype=type)) +
      geom_line() +
      labs(title = paste('Std Normal vs t with', df, 'degrees of freedom'))
    
    # actually print the nice graph we made
    print(myplot) 
```
    
a. Use a `for` loop to create similar graphs for degrees of freedom $2,3,4,\dots,29,30$.  I actually don't want all 30 graphs in the knitted pdf. Try this out, but we will alter the code in part b to get the plots we want in the knitted pdf. 
    
b. In retrospect, perhaps we didn't need to produce all of those. Rewrite your loop so that we only produce graphs for $\left\{ 2,3,5,10,15,20,25,30\right\}$ degrees of freedom. Hint: you can just modify the vector in the `for` statement to include the desired degrees of freedom. Show all 8 plots in your knitted pdf.

## Problem 2: Simulation

It is very common to run simulation studies to estimate probabilities that are difficult to work out. In this exercise we will investigate a gamblingquestion that sparked much of the fundamental mathematics behind the study of [probability](http://homepages.wmich.edu/~mackey/Teaching/145/probHist.html).
The game is to roll a pair of 6-sided dice 24 times. If a "double-sixes" comes
up on any of the 24 rolls, the player wins. What is the probability of winning?
    
a. We can simulate rolling two 6-sided dice using the `sample()` function with the `replace=TRUE` option. Read the help file on `sample()` to see how to sample from the numbers $1,2,\dots,6$. Sum those two die rolls and save it as `throw`.

b. Write a `for{}` loop that wraps your code from part (a) and then tests if any of the throws of dice summed to 12. Read the help files for the functions `any()` and `all()`. Your code should look something like the following:

```{r, eval=FALSE}
        throws <- NULL
        for( i in 1:24 ){
          throws[i] <- ??????  # Your part (a) code goes here
        }
        game <- any( throws == 12 ) # Gives a TRUE/FALSE value
```

c. Wrap all of your code from part (b) in *another* `for(){}` loop that you run10,000 times. Save the result of each game in a `games` vector that will have 10,000 elements that are either TRUE/FALSE depending on the outcome of that game. You'll need initialize the `games` vector to NULL and modify your part (b) code to save the result into some location in the vector `games`.

d. Finally, calculate win percentage by taking the average of the `games` vector.

## Problem 3

In the package `Lock5Data` there is a dataset `FloridaLakes` that contains water sample measurements from 53 lakes in Florida. Produce a bar graph that shows the `Lake` and `AvgMercury` variables, and make sure that the lakes are ordered by Average Mercury content. _To fix the issue of lake labels being squished together, you could rotate the labels using a `+ theme(axis.text.x = element_text(angle = 90, hjust = 1))` or flipping the coordinate axes. However, you'll likely want to resize the figure so the labels aren't so squished. the chunk options `fig.height` and `fig.width` control the size of the resulting figure._


## Problem 4

In the package `Lock5Data`, there is a dataset `FootballBrain` that has brain measurements for 75 individuals. The `Group` variable has three levels: `Control` is somebody that did not play football, `FBNoConcuss` is a football player with no history of concussions, or `FBConcuss` which is a football player with concussion history. The variable `Cogniton` measures their testing composite reaction time score. Change the `Group` labels to something that would make sense to a reader and create a box-plot graph of the groups vs cognition. *Because there is no data for the `Control` group, don't show it on your graph. Also notice that the original data set column name misspells "cognition".*

## Problem 5

In the package `Lock5Data`, there is a dataset `RestaurantTips` that gives tip data from the restaurant First Crush Bistro in Potsdam, NY. Make all necessary modifications to the `Day` variable so that subsequent graphs and tables are labeled in a professional manner. Do a similar modification to the `Credit` variable. Graph the `Bill` versus the `PctTip` for each `Day` of the week where we use `Day` as the variable to facet_grid or facet_wrap on.  Please make sure the days are in the order Sunday, Monday, Tuesday, ... Also include information about if the bill was paid via credit card or cash.




